Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-04 paper/paper/SPIE/article.tex
[WIP]

\section*{Purpose}
Random matrix theory (RMT) is an increasingly useful tool for understanding large, complex systems.
Prior studies have examined functional Magnetic Resonance Imaging (fMRI) scans using tools from
RMT, with some success. However, RMT computations are complex, and sensitive to numerous analytic
choices, and previous investigations have been mostly explanatory / descriptive. We systematically
investigate the usefulness of RMT in a variety of fMRI datasets in a predictive framework.

\section*{Approach}
We develop open-source software to efficiently compute RMT features from fMRI images, and examine
the cross-validated predictive potential of eigenvalue and RMT-based features (``eigenfeatures'') used
with typical classificatory machine-learning algorithms. We systematically vary pre-processing extent,
normalization procedure, RMT unfolding procedure, and classifiers to examine the utility of these
eigenfeatures across various datasets.

\section*{Results}
We find eiegenfeatures to be broadly predictive in a wide variety of contexts, and that RMT-specific
features like the spectral rigidity and level number variance have considerable potential in
feature-engineering for fMRI. On median, we find ... However, we also find the utility of the
extracted eigenfeatures to depend strongly and complexly on various analytic decisions.

Also

\section*{Conclusions}
Consider using RMT.

>>>>

\section*{Purpose}
Random matrix theory (RMT) is an increasingly useful tool for understanding
large, complex systems. Prior studies have examined functional Magnetic
Resonance Imaging (fMRI) scans using tools from RMT, with some success.
However, RMT computations are highly sensitive to a number of analytic choices,
and the robustness of findings involving RMT remain in question. We
systematically investigate the usefulness of RMT on a wide variety of fMRI
datasets using a rigorous predictive framework.

\section*{Approach}
We develop open-source software to efficiently compute RMT features from fMRI
images, and examine the cross-validated predictive potential of eigenvalue and
RMT-based features (``eigenfeatures'') with classic machine-learning
classifiers. We systematically vary pre-processing extent, normalization
procedures, RMT unfolding procedures, and feature selection, and compare the
impact of these analytic choices on the distributions of cross-validated
prediction performance for each combination of dataset binary classification
task, classifier, and feature. To deal with class imbalance, we use the
area under the receiver operating characteristic (AUROC) as the main
performance metric.

\section*{Results}
% df.groupby(["subgroup", "gross_feature"])["auroc"].median()
Across all classification tasks and analytic choices, we find eiegenfeatures to
more often have predictive utility than not (82.4\% of median AUROCs > 0.5;
median AUROC range across classification tasks 0.47 - 0.64). Simple baseline
reductions on source timeseries, by contrast, were less useful (58.8\% of
median AUROCs > 0.5, median AUROC range across classification tasks 0.42 -
0.62). Additionally, eigenfeature AUROC distributions were overall much more
right-tailed than baseline features suggesting greater predictive potential.
However, performance distributions were wide, and often significantly affected by
analytic choices.

\section*{Conclusions}
Eigenfeatures clearly have potential for understanding and reducing fMRI
functional connectivity. However, the utility of these features is strongly
dependent on analytic decisions, suggesting considerable caution when
interprting past and future studies applying RMT.
